---
title: "Homework_03"
author: "Stephen Bond"
date: "05/28/2026"
format: pdf
---

```{r}
#| label: Set Up
#| message: False
# read in tidyverse package
library(tidyverse)
# read in here package
library(here)
# read in janitor package
library(janitor)
# read in readxl package
library(readxl)
# read in ggeffects package
library(ggeffects)
# read in and make kelp data an object
kelp <- read_csv(here("data", "temp-kelp.csv"))
# read in and make personal data an object
dinner <- read_csv(here("data","Personal_Data - Sheet1.csv"))
```

# Problem 1. Giant kelp fronds

a. An Appropriate Test

The two appropriate tests that can be run to determine the strength of the relationship between temperature and giant kelp frond elongation rate are Pearson's correlation and Spearman rank correlation because they both answer the question of how two variables are related together and provide information on direction and strength of this relationship. Pearson's correlation test is a parametric version of a correlation test and it assumes that there is a linear relationship between variables, variables are continous, variables are normally distributed, and there are independent observations. On the other hand, Spearman rank correlation is the non-parametric version of this test that assumes a monotonic relationship between variables and that observations are independent.

b. Create a Visualization

```{r}
#|label: Visualizating Relationship between Variables
# base layer ggplot
ggplot(data = kelp,
       # assign x and y values
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # second layer: geom_point
  geom_point(
    # change color of points
    color = "darkblue", 
    # change shape and size of points
    alpha = 0.5, size = 3, shape = 20) +
  # label x- and y-axes
  labs(x = "Temperature (\u00B0C)",
       y = "Kelp Elongation Rate (cm/day)") +
  # change theme of plot
  theme_classic()
```

c. Check your Assumptions and Run your Test

###Check Linearity
```{r}
#|label: Visualizating Relationship between Variables
# base layer ggplot
ggplot(data = kelp,
       # assign x and y values
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # second layer: geom_point
  geom_point(
    # change color of points
    color = "darkblue", 
    # change shape and size of points
    alpha = 0.5, size = 3, shape = 20) +
  # label x- and y-axes
  labs(x = "Temperature (\u00B0C)",
       y = "Kelp Elongation Rate (cm/day)") +
  # change theme of plot
  theme_classic()
```

###Check Normality of Temperature 
```{r}
#| Label: QQ-Plot for Temperature
# base layer: ggplot
ggplot(data = kelp,
       # set sample to temperature
       aes(sample = temp_c)) +
# first layer: qq plot
geom_qq(color = "blue")+
# second layer qq line
geom_qq_line(color = "blue") +
# remove legend
theme(legend.position = "none")+
# change theme of plot
theme_minimal()
```

###Check Normality of Kelp Elongation Rate
```{r}
#| label: QQ-Plot for Kelp Elongation Rate 
# base layer: ggplot
ggplot(data = kelp,
       # set sample to kelp elongation rate
       aes(sample = kelp_elong)) +
# first layer: qq plot
geom_qq(color = "blue")+
# second layer qq line
geom_qq_line(color = "blue") +
# remove legend
theme(legend.position = "none")+
# change theme of plot
theme_minimal()
```


To determine whether to use Pearson's or Spearman Rank Correlation I had to check whether these variables had a linear or monotonic relationship and whether these variables have a normal distribution. I checked the general relationship between the two variables by creating a scatter plot, and based on initial gut check and the visual trend kelp elongation rate decreasing as temperature increases I concluded the relationship was linear. Additionally, to check that both these variables have normal distributions I create individual qq plots, and since both plots follow the generated diagonal line I concluded the distributions of temperature and kelp elongation rate were normal.

```{r}
#| label: Pearsons Correlation

cor.test(kelp$kelp_elong, # kelp elongation rat as first variable
         kelp$temp_c, # temperature as second variable
         method="pearson") # specify Pearson's correlation
```

d. Results Communication

To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used the Pearson's correlation test. I utilized this test because there was a seemingly linear relationship between the two variables, both variables are continous measurements with normal distributions, and there observations are independent. We found a strong negative relationship between temperature and giant kelp frond elongation rate (Pearson's r = -0.69, t(30) = -5.19, p << 0.001, $\alpha$ = 0,.05).

e. Test Implications

Our analysis found a strong negative relationship between ocean temperature (\u00B0C) and giant kelp frond elongation rate (cm^2^/day). This suggests that as ocean temperatures rise the rate of giant kelp frond growth decreases drastically, and while other factors might impact growth rates this strong correlation indicates that there will be miniscule impacts. In order to maximize kelp growth and general health we should aim to grow giant kelp in colder waters and to ensure the longevity of their health we must take actions to mitigate ocean temperature rising.

f. Double Check your own Work

```{r}
#| Label: Running alternative test
#| warnings: FALSE

# start correlation test
cor.test(kelp$temp_c, # set first variable to temperature
         kelp$kelp_elong, # set second variable to kelp elongation
         method = "spearman") # specify Spearman's Rank Correlation
```

The Spearman's Rank Correlation test would have yielded the same results as the Pearson's Correlation test. This would have lead me to makking the same decision regarding the null hypothesis, which would be to reject the null hypothesis because the variables temperature (\u00B0C) and kelp elongation rate (cm^2^/day) are related. We found a strong negative relationship between ocean temperature (\u00B0C) and kelp elongation rate (cm^2^/day)(Spearman $\rho$ = -0.69, S = 9216, p >> 0.01, $\alpha$ = 0.05)

# Problem 2. Personal Data

```{r}
#| label: Cleaning Data
# store dinner data as cooking_clean
cooking_clean <- dinner |>
  # clean names, making lowercase and underscore
  clean_names() |>
  # making order of days
  mutate(day_of_the_week_sun_mon_tues_wed_thur_fri_sat = 
           as_factor(day_of_the_week_sun_mon_tues_wed_thur_fri_sat),
         day_of_the_week_sun_mon_tues_wed_thur_fri_sat = 
           fct_relevel(day_of_the_week_sun_mon_tues_wed_thur_fri_sat,
                       "Sunday",
                       "Monday",
                       "Tuesday",
                       "Wednesday",
                       "Thursday",
                       "Friday",
                       "Saturday")) |>
  # name treatments
  mutate(cooked_at_home_y_n = as.numeric(case_when(
    # mutate yes to "1"
    cooked_at_home_y_n == "Yes" ~ "1",
    # mutate "No" into "0"
    cooked_at_home_y_n == "No" ~ "0")))
```

```{r}
#| label: Where I Cook Based on Day of the Week
# base layer: ggplot
ggplot(data = cooking_clean,
       # day of the week on the x-axis, binary cooked at home data on the y-axis)
       mapping = aes(x = day_of_the_week_sun_mon_tues_wed_thur_fri_sat,
                     fill = as.factor(cooked_at_home_y_n),
                     # color by whether I cooked at home or not
                     color = as.factor(cooked_at_home_y_n))) +
  # first layer: Bar graph
  geom_bar() +
  # relabelling x- and y-axes
  labs(x = "Day of the Week",
       y = "Count",
       # add title
       title = "Home Cooked Dinner Depending on Day of the Week",
       # add subtitle
       subtitle = "2026-05-28") +
  # manually change colors
  scale_color_manual(values = c("1"= "pink",
                                "0" = "black"),
                     guide = "none") +
  scale_fill_manual(values = c("1" = "pink",
                               "0" = "black" ),
                    labels = c("0" = "Ate Out",
                               "1" = "Cooked at Home"),
                    name = "Where I had Dinner") +
   # change theme from default
   theme_minimal() +
  # readjust days of the week for visibility
  theme(axis.text.x = element_text(angle = 45, hjust =1))+
  # cleaning grid lines and background
  theme(panel.grid = element_blank(),
        panel.background = element_blank())
```

```{r}
# creating model for cooking_clean data
cooking_model <- glm(cooked_at_home_y_n ~ time_worked_hours,
             data = cooking_clean,
             family = "binomial")
# creating new object called cooking_preds
cooking_preds <- ggpredict(
  cooking_model, # model object
  terms = "time_worked_hours" # predictor time worked
)
```

```{r}
#| label: Cooking at Home based on Hours Worked
# base layer: ggplot
ggplot(data = cooking_clean,
       # hours worked on the x-axis, binary cooked at home data on the y-axis
       # coloring by treatment
       mapping = aes( x = time_worked_hours,
                      y = cooked_at_home_y_n,
                      color = as.factor(cooked_at_home_y_n))) +
  # relabelling x- and y-axes
  labs(x = "Time Worked (Hours)",
       y = "Cooked at Home Y=1, N=0",
       # add color based on where dinner was cooked
       color = "cooked_at_home_y_n",
       # add title
       title = "Relationship between Hours Worked and my Cooking Habits",
       # add subtitle
       subtitle = "2026-05-28") +
  # manually change colors
  scale_color_manual(values = c("1" = "firebrick",
                                "0" = "darkblue" )) +
  # change theme from default
  theme_minimal()+
  # removing legend
  theme(legend.position = "none") +
  # second layer: jitter points horizontally
  geom_jitter(shape = 21, width = 0.25, height = 0) +
  # third layer: line representing mdoel prediction
  # using prediction data grame
  geom_line(data = cooking_preds,
            # set axis values
            aes(x = x,
                y = predicted),
            # alter line width
            linewidth = 1,
            inherit.aes = FALSE) +
  # cleaning grid lines and background
  theme(panel.grid = element_blank(),
        panel.background = element_blank())
```


