---
title: "Homework_03"
author: "Stephen Bond"
date: "05/28/2026"
format: pdf
---

```{r}
#| label: Set Up
#| message: False
# read in tidyverse package
library(tidyverse)
# read in here package
library(here)
# read in janitor package
library(janitor)
# read in readxl package
library(readxl)
# read in and make kelp data an object
kelp <- read_csv(here("data", "temp-kelp.csv"))
```

# Problem 1. Giant kelp fronds

a. An Appropriate Test

The two appropriate tests that can be run to determine the strength of the relationship between temperature and giant kelp frond elongation rate are Pearson's correlation and Spearman rank correlation because they both answer the question of how two variables are related together and provide information on direction and strength of this relationship. Pearson's correlation test is a parametric version of a correlation test and it assumes that there is a linear relationship between variables, variables are continous, variables are normally distributed, and there are independent observations. On the other hand, Spearman rank correlation is the non-parametric version of this test that assumes a monotonic relationship between variables and that observations are independent.

b. Create a Visualization

```{r}
#|label: Visualizating Relationship between Variables
# base layer ggplot
ggplot(data = kelp,
       # assign x and y values
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # second layer: geom_point
  geom_point(
    # change color of points
    color = "darkblue", 
    # change shape and size of points
    alpha = 0.5, size = 3, shape = 20) +
  # label x- and y-axes
  labs(x = "Temperature (\u00B0C)",
       y = "Kelp Elongation Rate (cm/day)") +
  # change theme of plot
  theme_classic()
```






